Thesis Title

### 2022-03-29 chapters/chapter_4.tex
update den-im paper

\section{Keywords}
dengue virus, surveillance, metagenomics, reproducibility, workflow, containerization, scalability

@@

CDS, coding sequence; DENV, dengue virus; HPC, high-performance computing; HTS, high-throughput sequencing; NCR, non-coding region; QC, quality control; RT-PCT, reverse transcription polymerase chain reaction.

@@

>>>>

\subsection{Keywords}
dengue virus, surveillance, metagenomics, reproducibility, workflow, containerization, scalability

@@

\ac{CDS}; \ac{DENV}; \ac{HPC}; \ac{HTS}; NCR, non-coding region; QC, quality control; RT-PCT, reverse transcription polymerase chain reaction.

@@

The viral genome of ~11,000 nucleotides, consists of a CDS of approximately 10.2 Kb that is translated into a single polyprotein encoding three structural proteins (capsid - C, premembrane - prM, envelope - E) and seven non-structural proteins (NS1, NS2A, NS2B, NS3, NS4A, NS4B and NS5). Additionally, the genome contains two Non-Coding Regions (NCRs) at their 5’ and 3’ ends \citep{leitmeyer_dengue_1999}.

DENV can be classified into four serotypes (1, 2, 3 and 4), differing from each other from 25\% to 40\% at the amino acid level. They are further classified into genotypes that vary by up to 3\% at the amino acid level \citep{diamond_molecular_2015}. The DENV-1 serotype comprises five genotypes (I-V), DENV-2 groups six (I-VI, also named American, Cosmopolitan, Asian-American, Asian II, Asian I and Sylvatic), DENV-3 four (I-III and V), and DENV-4 also four (I-IV).

The implementation of a surveillance system relying on HTS technologies allows the simultaneous identification and surveillance of DENV cases. Due to the high sensitivity of these technologies, previous studies showed that viral sequences can be directly obtained from patient sera using a shotgun metagenomics approach \citep{yozwiak_virus_2012}. Alternatively, HTS can be used in a targeted metagenomics approach in which a PCR step is used to pre-amplify viral sequences before sequencing. In recent years, HTS has been successfully used as a tool for identification of DENV directly from clinical samples \citep{yozwiak_virus_2012, lee_clinical_2017}. This also allows the rapid identification of the serotype and genotype important for disease management as the genotype may be associated with disease outcome \citep{fatima_serotype_2011}.

Several initiatives aim to facilitate the identification of the DENV serotype and genotype from HTS data. The Genome Detective project (https://www.genomedetective.com/) offers an online Dengue Typing Tool (https://www.genomedetective.com/app/typingtool/dengue/) \citep{fonseca_computational_2019} relying on BLAST and phylogenetic methods in order to identify the closest serotype and genotype, but it requires as input assembled genomes in FASTA format. The same project also offers the Genome Detective Typing Tool (https://www.genomedetective.com/app/typingtool/virus/) \citep{vilsker_genome_2019} identifying viruses present in a sample. Additionally, there are several tools available for viral read identification and assembly, such as VIP \citep{li_vip_2016}, virusTAP \citep{yamashita_virustap_2016} and drVM \citep{lin_drvm_2017}, but none performs genotyping of the identified reads.

We developed DEN-IM as a ready-to-use, one-stop, reproducible bioinformatic analysis workflow for the processing and phylogenetic analysis of DENV using paired-end raw HTS data. DEN-IM is implemented in Nextflow \citep{di_tommaso_nextflow_2017}, a workflow manager software that uses Docker (https://www.docker.com) containers with pre-installed software for all the workflow tools. The DEN-IM workflow, as well as parameters and documentation, are available at \url{https://github.com/B-UMMI/DEN-IM}.

\section{The DEN-IM Workflow}
